---
title: "An Analysis of Test Scores Against Various Social and Economic Factors"
author: "Allison Zhang, Kobe Dela Cruz, Nishant Balepur"
date: "11/30/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

The purpose of our project is to effectively predict the test scores of high schoolers in 1986 based on a large number of predictor variables, some of which are gender, ethnicity, family income, etc. 


Our ultimate goal for this project is to reason through the different predictor variables to find the model that can best predict high school test scores.


## Dataset

We will be using the College Distance dataset from the AER package that is found on this following site: https://rdrr.io/cran/AER/man/CollegeDistance.html

The data was obtained by the Department of Education, and contains many different social and economic variables, including: gender, ethnicity, whether or not the mother/father went to college, if the family owns their home, county, unemployment rate, and more.

We first load 'college_distance.csv' into R:

```{r}
library(readr)
library(knitr)
collegedistance = read_csv("CollegeDistance.csv")
```
This dataset has a total of 4739 observations of 15 variables (14 predictors and 1 response). The said variables are:

- gender: a factor indicating gender
- ethnicity: factor indicating ethnicity (African-American, Hispanic or other)
- score: base year composite test score
- fcollege: factor. Is the father a college graduate?
- mcollege: factor. Is the mother a college graduate?
- home: factor. Does the family own their home?
- urban: factor. Is the school in an urban area?
- unemp: country unemployment rate in 1980
- wage: state hourly wage in manufacturing in 1980
- distance: distance from 4-year college (in 10 miles)
- tuition: average state 4-year college tuition (in 1000 USD)
- education: number of years of education
- income: factor. Is the family income above 25,000 USD per year?
- region: factor indicating region (West or other)


The dataset meets all of the set criteria for the project. Now lets look for missing values for our next step in data cleaning.
```{r}
summary(collegedistance)
```


## Data Cleaning

## Method Exploration

## Plotting

## Model Results